The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-08-29 paper.tex
contributions from Mittweida

%\pretitle{}
\title{The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes}
\runtitle{Anthropological Notation Ontology}
%\subtitle{}

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
}
\author[B,C]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
\address[C]{equal contribution}
\end{aug}

@@

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, an application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

Digitization offers several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
Moreover it allows for linking data of different study samples across geographical boundaries.
Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.

@@

ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures. In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
ANNO is integrated into AnthroWorks3D (AW3D), an application for generating and analyzing 3D-models of human skeletal remains.
ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.

@@

\subsection{Problems and Research Gaps}

@@

To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
These definitions should be presented within an easily understandable ontology.

@@

\begin{enumerate}
\item We use the General Formal Ontology (GO)~\citep{gfo} as top-level ontology of ANNO.
We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},

@@

\subsection{Phenotype}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/phenotype.png}
\caption{
Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
% yes, walker method is published in buikstra
Method after Walker~\cite{datacollection}.
Image based on \cite{datacollection}.
}\label{fig:phenotype}
\end{figure}

@@

In the case of ANNO, the phenotype can be derived using functions comprising obtained measurements.
Discriminant functions, e.g based on the Cranium, allow for the assignment to sex while estimation of stature is attained by using linear regressions, e.g. based on Femur or Tibia length measurements.
RDF is not optimized for mathematical formulas so we model those as literals.

@@

In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
%
%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
%\subsections{General considerations}
%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
%
%Anatomical Landmarks (Bone Parts)  (Selection Criteria, scope and process,  properties) 
%Osteometric Landmarks (Anatomical Points) 
%Measurements (Anatomical line)  
%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
%Phenotypes 
%Development of the ontology is based on the following selection criteria:
%The selection criteria are as follows:
For the cranium, the selection is reduced because the number of all ossa cranii is too high.
Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
Those landmarks included in the definitions of others were also to be defined.
For bones that lie on the median (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
For osteometric landmarks, all those already established in the core literature are to be used.
Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
%In addition, it also had to be integrable into \aw{}.
The functions chosen were those with diagnostic value.
These were discriminant functions for sex determination and regression functions for body height and body weight estimation.

%\subsection{Process for creating definitions and measurements}
For the definitions and measurements, a minimum amount of English- and mostly German-language literature was compared in order to develop the definitions from their information.
Notably, Latin or ancient Greek terms were missing from the English literature and the FMA.
%The FMA also contains only English terms.
%For this reason, these were also searched with their synonyms.
Overall, the name of the landmark or survey route is noted in English, German, and Latin in singular and plural form, its synonyms in the three languages, the FMA and TA ID, and any information on function and delineation.
This requires a positional description, such as the Punctum superioris capitis femoris as the superior located point of the Caput femoris.
%The examination of the Latin grammar was relevant for this.
For the measurement sections, in addition to the name of the measurement section, the type of measurement (e.g., distance measurement) and the measurement instrument were also recorded.
The subsequent visual definitions were made in the different anatomical views and represented area markings for the anatomical landmarks and point markings for the osteometric landmarks.

%\subsection{Functions}
The functions are divided into discriminant functions for sex determination and regress functions for body height and body weight estimation.
The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific to this population.
Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female.
ANNOds covers functions for at least one European, African, American, and Asian ethnicity or population.
%Names are assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender are noted.
%For the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification are important; for the regression functions, the error interval is important.
%
%While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
%ANNO links to concepts of the FMA when they exist, but structures them in a different way.
%
%\subsection{Sex determination}
% auto translated from the project report, use as basis:
%
%\subsection{Regress functions}
%Regress functions for body height and body weight estimation.
%Names were assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
%In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
%
%Standardised description of position 
%In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered. Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet. In the latter case, the OWL file can be generated with a custom OWL generator.

@@

The ontology is interlinked with the TA and FMA and provides transparency by including all sources used to generate its content.
Moreover, it provides a method for conceptualizing the ontology and generating its content.
The ontology replaces an old, hard-coded format in \aw{}, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and improves interoperability and customization.
Also, ANNO comes with extensive documentation of the process and available resources.
Hence, foundations have been laid for its use and further development and management of ANNO.

@@

While ANNO serves as a foundation for describing anatomical terms, additional work is required to comprehensively cover the domain.
For instance, it only covers standardized normal adult anatomy.
Moreover, in order to remain permanently suitable for practical usage, the ontology must be continuously reviewed, updated, enriched or adapted to further needs.
Moreover, ANNO presents all necessary prerequisites to being extended to the other aspects of anthropological work with human skeletal remains.

>>>>

%\pretitle{}
\title{The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes}
\runtitle{Anthropological Notation Ontology}
%\subtitle{}

% For one author:
%\author{\inits{N.}\fnms{Name1} \snm{Surname1}\ead[label=e1]{first@somewhere.com}}
%\address{Department first, \orgname{University or Company name},
%Abbreviate US states, \cny{Country}\printead[presep={\\}]{e1}}
% Höffner, Heuschkel, Schmiedel, Penne, Fritzsch, Ludwig, Mohaupt, Labudde, Uciteli
% TODO: overleaf
% Two or more authors:
\begin{aug}
\author[B,C]{\inits{M.}\fnms{Marie} \snm{Heuschkel}\ead[label=e2]{marie.heuschkel@hs-mittweida.de}}
\author[A,C]{\inits{K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}
}
\author[B]{\inits{F.}\fnms{Fabian} \snm{Schmiedel}\ead[label=e3]{fabian.schmiedel@hs-mittweida.de}}
\author[B]{\inits{D.}\fnms{Dirk} \snm{Labudde}\ead[label=e10]{dirk.labudde@hs-mittweida.de }}
\author[A]{\inits{A.}\fnms{Alexandr} \snm{Uciteli}\ead[label=e11]{alexander.uciteli@imise.uni-leipzig.de}}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE), \orgname{Leipzig University},
Saxony, \cny{Germany}\printead[presep={\\}]{e1,e11}}
\address[B]{\orgname{Hochschule Mittweida},
Saxony, \cny{Germany}\printead[presep={\\}]{e2,e3,e10}}
\address[C]{equal contribution}
\end{aug}

@@

\begin{abstract}
The Anthropological Notation Ontology (ANNO) allows the systematic and standardized classification of recovered bone finds into the skeletal system, the description of the skeletal pieces, and the definition of functions for the derivation of different phenotypes of humans in forensic and historical anthropology.
ANNO consists of two components:
ANNOdc, a domain-core ontology providing core entities such as basic anatomical categories, and ANNOds, a domain-specific ontology used for annotating structures of the human skeleton.
ANNO is integrated into AnthroWorks3D, a photogrammetry pipeline and application for the creation and analysis of 3D-models of human skeletal remains.
The integration is based on the three-ontology method with the General Formal Ontology as the top level ontology, ANNOdc as the task ontology and ANNOds as the domain ontology.
Thus, AnthroWorks3D only needs to implement access to the entities (classes and properties) of the task ontology, whereas the entities of the corresponding domain ontology are processed dynamically.
ANNO supports the analysis of skeletal and bone finds in forensic and historical anthropology, facilitating the standardization of data annotation and ensuring accurate preservation of information for posterity.
\end{abstract}

@@

Digitization and digitalization offer several advantages to anthropology, including remote work, without the effects of wear and tear on the physical skeletal specimens.
Current problems of access can be resolved, as examinations can be conducted even if the skeletal material from individuals or collectives are not present at the institution or may have already been reburied, which facilitates interdisciplinary collaborative research.
Moreover it allows for linking data of different study samples across geographical boundaries.
Data sets consolidated this way and formalized through the Anthropological Notation Ontology (ANNO) presented here allow for efficient data analysis techniques, such as data and text mining that promote deeper insights.

@@

ANNO consists of two components: ANNOdc, a domain-core ontology providing core entities such as basic anatomical concepts and classifications, and ANNOds, a domain-specific ontology employed for annotating human skeletal structures. In its current version, the ontology refers to standardized normal adult anatomy, excluding developmental aspects, variations, pathologies and other related factors.
ANNO is integrated into AnthroWorks3D (AW3D), a photogrammetry pipeline and application for generating and analyzing 3D-models of human skeletal remains.
ANNO is developed at \url{https://github.com/annosaxfdm} and published at \url{https://annosaxfdm.de/ontology/} using the RickView~\citep{rickview} browser.

@@

\subsection{Current Research Gaps}

@@

To address these issues, it is necessary to develop unified, textual, visual, and detailed descriptive definitions for anatomical concepts and instances.
These definitions should be presented within an easily understandable ontology, with ANNO contributing to filling this specific niche.

@@

\begin{enumerate}
\item We use the General Formal Ontology (GFO)~\citep{gfo} as top-level ontology of ANNO.
We refer specifically to the categories \enquote{Material object}, \enquote{Attributive}, \enquote{Relator} as well as the spatial entities of GFO, see \cref{fig:gfo}. %\cref{sec:core},

@@

\subsection{Phenotype}
\begin{figure}[h]
\includegraphics[width=\textwidth]{img/phenotype.png}
\caption{
Exemplary assessment of anatomical properties for the derivation of the phenotype sex, where five cranial bone parts are evaluated for their level of morphological expression.
% yes, walker method is published in buikstra
Method after Walker~\cite{datacollection}.
Image taken unmodified from \cite{datacollection}.
}\label{fig:phenotype}
\end{figure}

@@

In the case of ANNO, the phenotype can be derived using functions comprising obtained measurements.
Discriminant functions, e.g. based on the Cranium, allow for the assignment to sex while estimation of stature is attained by using linear regressions, e.g. based on Femur or Tibia length measurements.
RDF is not optimized for mathematical formulas so we model those as literals.

@@

In addition, the objective is to initiate the process of data entry, encompassing selected bones of the skeleton.
%
%\textbf{UNFINISHED DRAFT SECTION, TO BE HEAVILY REWORKED AND CUT}
%\subsections{General considerations}
%All landmarks are annotated with their name in singular and plural form, reference bone, synonyms, textual and visual definition, FMA and TA ID, and sources.
%
%Anatomical Landmarks (Bone Parts)  (Selection Criteria, scope and process,  properties) 
%Osteometric Landmarks (Anatomical Points) 
%Measurements (Anatomical line)  
%Relevant spatial anatomical objects (e.g., lines) used for phenotype determination functions Measurements were defined in terms of anatomical points/objects that bound them: consist of references to sections, the start point, midpoint, and end, and short definitions.
%Phenotypes 
%Development of the ontology is based on the following selection criteria:
%The selection criteria are as follows:
Given the overwhelming number of bone parts present in the cranium, the initial selection was narrowed down to a selection of representative and relevant parts.
Overall, however, the aim is to include those that are of anthropological relevance, i.e. that contribute to navigation, localization, and identification on the bone.
Those structures included in the definitions of others were also to be defined.
For bones that lie on the median sagittal plane (e.g., mandible or sternum), bilateral landmarks are defined, each with a side reference.
For osteometric landmarks, all those already established in the core literature are to be used.
Furthermore, those that are relevant for meaningful measurement distances and can be annotated in \aw{} should be used.
For the measurement distances, those should be selected that are established in the majority of the core literature as well as necessary for discriminant functions and functions for estimating height and weight.
%In addition, it also had to be integrable into \aw{}.
The functions chosen were those with diagnostic value.
These were discriminant functions for sex determination and regression functions for body height and body weight estimation.

%\subsection{Process for creating definitions and measurements}
For the definitions and measurements, a representative minimum amount of anthropological and anatomical English- and mostly German-language literature was compared in order to develop the definitions from their information.
Notably, Latin or latinized ancient Greek terms often missing in the English literature and the FMA were included.
%The FMA also contains only English terms.
%For this reason, these were also searched with their synonyms.
Overall, the name of the structure, measurement or point is noted in Latin in singular and plural forms, English, German, synonyms in all three languages, the FMA and TA ID, and any information on function and delineation.
This requires a positional description, such as the Punctum superioris capitis femoris as the superior located point of the Caput femoris.
%The examination of the Latin grammar was relevant for this.
For the measurements, in addition to the name of the measurement, the type (e.g., distance measurement) and the measurement instrument were also recorded.
The subsequent visual definitions were made in the different anatomical views marking the area of the anatomical structures and the position of the anatomical points.

%\subsection{Functions}
The functions are divided into discriminant functions for sex determination and regress functions for body height.
The sex of a specific individual within a population may be estimated using a function on skeletal measurements that is specific or similar to this population.
Based on a threshold value, skeletons are classified into male, probably male, indifferent, probably female, and female. The exact number of categories may vary depending on the particular method used.
ANNOds covers functions for at least one European, African, American, and Asian ethnicity or population.
%Names are assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender are noted.
%For the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification are important; for the regression functions, the error interval is important.
%
%While there are many textual sources of anthropological systematisation, they do not agree in all aspects and there is not a single, formally described, standard.
%ANNO links to concepts of the FMA when they exist, but structures them in a different way.
%
%\subsection{Sex determination}
% auto translated from the project report, use as basis:
%
%\subsection{Regress functions}
%Regress functions for body height and body weight estimation.
%Names were assigned by the number included in each study, the authors, and the year of publication.
%In addition, the function, reference population, aspect (e.g., discriminant function), and sample size with division by gender were noted.
%In addition, for the discriminant functions, the thresholds of sex assignment, classification accuracy, and misclassification were important; for the regression functions, the error interval was important.
%
%Standardised description of position 
%In addition to the conception and development of the ontology as well as of definition, data input was initiated as content of Cranium including the Mandibula as well as the teeth of the was entered. Entries can be added directly to the OWL-File using Protégé or they can continue to be entered directly into the spreadsheet. In the latter case, the OWL file can be generated with a custom OWL generator.

@@

The ontology is interlinked with the TA and FMA and provides transparency by including all sources used to generate its content.
Moreover, it provides a method for conceptualizing the ontology and generating its content.
The ontology replaces an old, hard-coded format in \aw{}, which saves development time, separates annotation file compatibility from software versions, eases annotation through hierarchical browsing and improves interoperability and customization.
Also, ANNO comes with extensive documentation of the process and available resources.
Hence, foundations have been laid for its use and further development and management.

@@

While ANNO serves as a foundation for describing anatomical terms, additional work is required to comprehensively cover the anthropological domain.
For instance, it only covers standardized normal adult anatomy.
Moreover, in order to remain permanently suitable for practical usage, the ontology must be continuously reviewed, updated, enriched or adapted to further needs.
Moreover, ANNO presents all necessary prerequisites to being extended to the other aspects of anthropological work with human skeletal remains.
